A family of accessibility measures derived from spatial interaction principles

### 2025-04-05 paper/manuscript.qmd
oh Hansen, you broke my heart

\<!-- I suggest removing 'using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020]' from the paragraph above --!\>

\<!-- the first two paragraphs set the context / expectaction for the paper. In the current form, it the first two paragraphs seem to be more focused on the importance of accessibility for planning, and the challenges about adopting accessibility into planning practice. However, the paper is very much a theoretical and methodological contribution. the first paragraphs should reflect this, drawing attention to the common origin of spatial interaction models and accessibility , and how they diverged over time and we got lost along the way . Some pepople might not have a clear understanding of how the two concepts connect or how exactly they differ. this could be briefly explained --!\>

@@

\<!-- the paragraph above is really good --!\>

@@

\<!-- Most readers will not be familiar with your previous papers, so they will not understand what you mean with 'single constraint' or 'balancing factors'. These ideas need to be explained in this paper. I suggest the text 'There have also been more recent.....' in the paragraph is moved to the literature review section, where we have proper space to explain those ideas and how the ideas proposed in this paper expand on then --!\>

@@

\<!--

when you say that the absence of empirical constraints has led to interpretation issues in accessibility analysis, and think the example should be stronger that simply saying it leads to the 'fuzziness of insights' . perhaps mention the problem that we loose track of measurement units, that comparability between indicators, across cities and transport modes also become compromised but most of us don't even notice it, and this should be explained in the next sections

--!\>

@@

To be certain, the expression above, a proportionality, is one of the most famous in all of science. In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly *proportional* to the product of their masses $M_i$ and $M_j$, and inversely *proportional* to the square of the distance between them $D_{ij}$. Direct proportionality means that as the product of the masses increases, so does the force. Likewise, inverse proportionality means that as the distance increases, the force decreases. @eq-phys-grav-prop, however, does not quantify the magnitude of the force. To do so, an empirical constant is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses. In other words, @eq-phys-grav-prop needs to be *constrained* using empirical data. Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically calibrated proportionality constant:

@@

```{=html}
<!-- 
SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s. A number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
-->
```

@@

After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948]. This may relate to academic backgrounds; where Ravenstein, Reilly, and Zipf were social scientists, Stewart was a physicist. Besides awareness of preceding research (he cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:

@@

In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows: $$
V_1 = G\frac{M_2}{d_{12}}
$$ {#eq-stewart-population-potential}

@@

<!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up. Is now cited as the father of accessibility. However, Hansen work was applied, and he's use of Stewart's Population Potential measure included two crucial omissions that afflict the literature to this day. The first omission is 1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it. The second omission is 2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient. In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed. -->

From @stewartDemographicGravitationEvidence1948, we arrive to 1959 and Walter G. Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959]. In his seminal paper, Hansen defined accessibility as "the potential of opportunities for interaction... a generalization of the population-over-distance relationship or *population potential* concept developed by @stewartDemographicGravitationEvidence1948" (p. 73). As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense. @hansen1959 focused on @stewartDemographicGravitationEvidence1948's *population potential* (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind. @hansen1959 recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns. @hansen1959 replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more accurately, a *potential of opportunities for interaction* as follows:

@@

Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished. There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about *directly* and *inversely proportional* relationships when discussing population, opportunities, and their separation in space. At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity".

After @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained implicit or forgotten, with no notable developments to explicitly determine it. In this way, $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i}  \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. Alas, without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated. For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta\cdot m}$ when $f(c_{ij}) = e^{\beta\cdot m}$. Hansen-style accessibility, therefore, is better thought of as an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility [@millerAccessibilityMeasurementApplication2018].

>>>>

<!-- I suggest removing 'using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020]' from the paragraph above -->

<!-- the first two paragraphs set the context / expectaction for the paper. In the current form, it the first two paragraphs seem to be more focused on the importance of accessibility for planning, and the challenges about adopting accessibility into planning practice. However, the paper is very much a theoretical and methodological contribution. the first paragraphs should reflect this, drawing attention to the common origin of spatial interaction models and accessibility , and how they diverged over time and we got lost along the way . Some pepople might not have a clear understanding of how the two concepts connect or how exactly they differ. this could be briefly explained -->

@@

<!--  the paragraph above is really good -->

@@

<!-- Most readers will not be familiar with your previous papers, so they will not understand what you mean with 'single constraint' or 'balancing factors'. These ideas need to be explained in this paper. I suggest the text 'There have also been more recent.....' in the paragraph is moved to the literature review section, where we have proper space to explain those ideas and how the ideas proposed in this paper expand on then -->

@@

<!-- when you say that the absence of empirical constraints has led to interpretation issues in accessibility analysis, and think the example should be stronger that simply saying it leads to the 'fuzziness of insights' . perhaps mention the problem that we loose track of measurement units, that comparability between indicators, across cities and transport modes also become compromised but most of us don't even notice it, and this should be explained in the next sections -->

@@

To be certain, the expression above, a proportionality, is one of the most famous in all of science. In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly *proportional* to the product of their masses $M_i$ and $M_j$, and inversely *proportional* to the square of the distance between them $D_{ij}$. Direct proportionality means that as the product of the masses increases, so does the force. Likewise, inverse proportionality means that as the distance increases, the force decreases. @eq-phys-grav-prop, however, does not quantify the magnitude of the force. To do so, an empirical constant, aka gravitational constant, is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses. In other words, @eq-phys-grav-prop needs to be *constrained* using empirical data. Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically calibrated proportionality constant:

@@

```{=html}
<!-- 
SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s. A number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
```

Since the 1880s to the 1940, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.

@@

After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948]. This may relate to academic backgrounds; where Stewart was a physicist while Ravenstein, Reilly, and Zipf were social scientists. Besides awareness of preceding research (he cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:

@@

<!-- what exactly did he mean by 'molecular weights' ? -->

In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows:

<!-- in plain language, what do you mean by 'the population potential of 2 with respect to 1' ? the potential number of people from location 2 that would visit location 1 ? but also, I'm not sure we need this equation 7 -->

\
$$
V_1 = G\frac{M_2}{d_{12}}
$$ {#eq-stewart-population-potential}

@@

<!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up and is now cited as the father of accessibility. However, Hansen work was applied, and he's use of Stewart's Population Potential measure included two crucial omissions that afflict the literature to this day. The first omission is 1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it. The second omission is 2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient. In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed. -->

From @stewartDemographicGravitationEvidence1948, we arrive to 1959 and Walter G. Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959]. In his seminal paper, Hansen defined accessibility as "the potential of opportunities for interaction... a generalization of the population-over-distance relationship or *population potential* concept developed by @stewartDemographicGravitationEvidence1948" (p. 73). As well as being a student of city and regional planning at the Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense. @hansen1959 focused on @stewartDemographicGravitationEvidence1948's *population potential* (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind. @hansen1959 recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns. @hansen1959 replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more accurately, a *potential of opportunities for interaction* as follows:

@@

However, Hansen' use of Stewart's *population potential* measure included one crucial omission that afflict the literature to this day. The omission is that that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished. This constant was not explicitly addressed in @hansen1959 and accessibility research continues to evolve without it. Since @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained implicit or forgotten, with no notable developments to explicitly determine it.

Paper, the omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which has fundamental implications for the interpretation, communication and comparability of accessibility analysis. At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity". In practice, the persistent omission of the costant in accessibility analyses  means that $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i}  \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. The direct consequence is that without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated. For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta\cdot m}$ when $f(c_{ij}) = e^{\beta\cdot m}$. This undermines the comparability of accessibility metrics with different decay functions, and renders their results difficult to understand and communicate. Hansen-style accessibility, therefore, is better thought of as an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility, but which has not palpable meaning [@millerAccessibilityMeasurementApplication2018].

<!-- jargon ? 'a suitable choice of other units can reduce it to unity' I think this could be explained in plain English ->
